---
title: obisrecipes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

> **Internal use only.** This package is developed for the
> [OBIS Secretariat](https://obis.org) and OBIS nodes. The API may change
> without notice as internal workflows evolve.

`obisrecipes` is an R package of reusable visualization and mapping functions
for [Ocean Biodiversity Information System (OBIS)](https://obis.org) data. It
covers three areas:

- **ggplot2 styles** — a light theme, a dark marine theme, colorblind-safe
  color scales, a horizontal colourbar guide, and an OBIS/IOC logo overlay.
- **Static maps** — ggplot2-based world basemaps, grid converters (geohash,
  H3, A5), and an occurrence layer for plotting species distributions.
- **Dynamic maps** — interactive WebGL maps via `mapgl`, a DeckGL H3 hexagon
  widget, and helpers for adding species data and map controls.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("iobis/obis-recipes")
```

## Quick start

### ggplot2 styles

```{r ggplot-example}
library(obisrecipes)
library(ggplot2)
library(robis)
suppressPackageStartupMessages(library(dplyr))
obis_load_fonts()

occ_acch <- occurrence(taxonid = 159580, startdate = "2010-01-01")

ggplot(occ_acch, aes(sst, depth)) +
    geom_point(size = 2.5, color = "#188eb6", alpha = .2) +
    labs(title = "Acanthurus chirurgus",
         subtitle = "Doctorfish",
         x = "SST (°C)", y = "Depth (m)") +
    theme_obis()
```

```{r ggplot-example2}
occ_acan <- occurrence(
    taxonid  = 125515,
    startdate = "2015-01-01",
    geometry = "POLYGON ((-81.5625 31.653381, -97.910156 30.448674,
                          -99.492188 21.125498, -92.460937 16.636192,
                           -83.847656 11.005904, -74.707031 7.188101,
                            -58.710938 7.536764, -52.910156 13.068777,
                             -61.171875 28.767659, -73.125 34.016242,
                              -81.5625 31.653381))"
)

occ_acan_agg <- occ_acan |>
    group_by(date_year, species) |>
    summarise(Records = n(), .groups = "drop") |>
    rename(Year = date_year, Species = species) |>
    filter(!is.na(Species))

ggplot(occ_acan_agg, aes(Year, Records, fill = Species)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_fill_obis_cat() +
    labs(title = "Acanthuridae",
         subtitle = "Surgeonfishes — Caribbean records since 2015",
         x = "Year", y = "Records") +
    theme_obis(legend_position = "bottom")
```

### Colorbars

```{r colorbar-light, fig.height = 4}
sst_caribbean_bo <- terra::rast(
    "https://erddap.bio-oracle.org/erddap/griddap/thetao_baseline_2000_2019_depthsurf.nc?thetao_mean%5B(2010-01-01T00:00:00Z):1:(2010-01-01T00:00:00Z)%5D%5B(0.025):1:(43.025)%5D%5B(-103.975):1:(-37.975)%5D"
)

sst_caribbean_bo <- terra::aggregate(sst_caribbean_bo, 4) |>
    as.data.frame(xy = TRUE)

ggplot(sst_caribbean_bo) +
    geom_raster(aes(x = x, y = y, fill = thetao_mean)) +
    scale_fill_obis_cont(palette = "thermal") +
    guides(fill = guide_colourbar_h(title = "Temperature (°C)",
                                    barwidth = grid::unit(14, "lines"))) +
    labs(x = NULL, y = NULL, title = "Sea surface temperature — 2010 average") +
    scale_x_continuous(expand = FALSE) +
    scale_y_continuous(expand = FALSE) +
    theme_obis(legend_position = "bottom") +
    theme(panel.grid = element_blank()) +
    coord_equal()
```

```{r colorbar-dark, fig.height = 4}
ggplot(sst_caribbean_bo) +
    geom_raster(aes(x = x, y = y, fill = thetao_mean)) +
    scale_fill_obis_cont(palette = "thermal") +
    guides(fill = guide_colourbar_h(title = "Temperature (°C)",
                                    barwidth = grid::unit(14, "lines"))) +
    labs(x = NULL, y = NULL, title = "Sea surface temperature — dark theme") +
    scale_x_continuous(expand = FALSE) +
    scale_y_continuous(expand = FALSE) +
    theme_obis_dark(legend_position = "bottom") +
    theme(panel.grid = element_blank()) +
    coord_equal()
```

### Static maps

```{r static-basemap, fig.height = 4}
grey_map_s()
```

Focused regional view with gridded occurrence data:

```{r static-species, fig.height = 4.5}
grey_map_s() |>
    add_species_data_s(
        grid_data     = occ_to_geohashgrid(occ_acch, grid_res = 3),
        limit_by_bbox = FALSE,
        plot_xlim     = c(-90, -34),
        plot_ylim     = c(-28, 43)
    )
```

### Dynamic maps

Interactive WebGL maps rendered with [mapgl](https://walker-data.com/mapgl/).
These are fully interactive — pan, zoom, and click to explore.

```{r dynamic-globe}
grey_map() |>
    add_species_data(taxonid = 126436, legend = TRUE)
```

```{r dynamic-mercator}
grey_map(projection = "mercator") |>
    add_species_data(taxonid = 126436, legend = TRUE)
```

The following examples use DuckDB to pull pre-gridded H3 data from the OBIS
S3 product and render it with the DeckGL H3 widget. Run them interactively
after installing the optional dependencies (`duckdb`, `h3jsr`).

```{r dynamic-h3, eval = FALSE}
library(duckdb)
library(h3jsr)

con <- dbConnect(duckdb())
df <- dbGetQuery(con, "
    SELECT records, cell
    FROM read_parquet('s3://obis-products/speciesgrids/h3_7/*')
    WHERE species = 'Minuca rapax'
")
df <- df |>
    mutate(h3 = h3jsr::get_parent(cell, 3)) |>
    group_by(h3) |>
    summarise(value = sum(records))

# Flat H3 layer
maplibre_h3(df, center = c(-50, -10), zoom = 2)

# Extruded 3-D columns
maplibre_h3(df, center = c(-50, -10), extruded = TRUE, elevation_scale = 3000)
```

## Vignettes

| Article | Topic |
|---------|-------|
| [ggplot2 styles](articles/ggplot-styles.html) | Themes, color scales, colourbar guide, logo overlay |
| [Static maps](articles/static-maps.html) | Basemaps, grid converters, offline and API-based plotting |
| [Dynamic maps](articles/dynamic-maps.html) | Interactive WebGL maps and the H3 widget |

---

Developed by the [OBIS Secretariat](https://obis.org) · Hosted under the
[Intergovernmental Oceanographic Commission (IOC)](https://www.ioc.unesco.org)
of UNESCO.
