---
title: "Dynamic Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(obisrecipes)
```

The dynamic-map functions produce interactive WebGL maps (via `mapgl`) that
render in the RStudio viewer, R Markdown documents, and Shiny apps. The
workflow mirrors the static-map approach:

1. Create a basemap with `blank_map()`, `grey_map()`, or `carto_map()`.
2. Add species occurrence data with `add_species_data()`.
3. Optionally add UI controls with `add_map_controls()`.

> **Note:** All examples in this vignette require an internet connection
> because the map tiles and OBIS occurrence data are fetched at render time.

## Basemaps

### Blank white basemap

```{r blank-globe}
blank_map()
```

Switch to a flat Mercator projection:

```{r blank-mercator}
blank_map(projection = "mercator")
```

### Grey OBIS-style basemap

`grey_map()` matches the style used on OBIS taxon pages: grey ocean, white
land, optional GEBCO bathymetry overlay.

```{r grey-default}
grey_map()
```

```{r grey-no-bath}
grey_map(add_bathymetry = FALSE)
```

### Carto basemaps

```{r carto-positron}
carto_map()
```

```{r carto-dark}
carto_map(carto_map = "dark-matter")
```

## Species occurrence layer

`add_species_data()` fetches OBIS occurrence tiles for a taxon and adds a
colour-scaled fill layer to the map.

Default species (taxonid 126436) on a blank map:

```{r occ-default}
blank_map() |> add_species_data()
```

Custom species with custom breaks and a legend:

```{r occ-custom}
grey_map() |>
    add_species_data(
        taxonid = 955271,
        breaks  = c(0, 5, 50, 500, 5000),
        legend  = TRUE
    )
```

Fit the map view to the species extent after adding the layer:

```{r occ-fit-bounds}
ext <- get_species_extent(taxonid = 955271)

blank_map(projection = "mercator") |>
    add_species_data(taxonid = 955271) |>
    fit_bounds(ext)
```

### Using OBIS palettes

Pass any named OBIS palette via the `palette` argument and `add_species_data()`
will call `obis_pal()` internally to generate one colour per break. This works
with all palettes from `scale_fill_obis_cont()` and `scale_color_obis_cat()`.

```{r occ-palette-thermal}
grey_map() |> add_species_data(taxonid = 126436, palette = "thermal")
```

```{r occ-palette-blues}
grey_map() |> add_species_data(taxonid = 126436, palette = "blues")
```

Pair fewer breaks with a diverging palette:

```{r occ-palette-rdbu}
blank_map() |>
    add_species_data(
        taxonid = 126436,
        breaks  = c(0, 10, 100, 1000),
        palette = "rdbu",
        legend  = TRUE
    )
```

`obis_pal()` can also be called directly to preview colours or pass them to
other `mapgl` layer functions:

```{r obis-pal-direct}
obis_pal("thermal", 5)
#> [1] "#042333" "#4a69a9" "#cde6c9" "#f0a120" "#b12010"

obis_pal("rdbu", 7)
obis_pal("blues", 4, direction = -1)  # reversed: dark → light
```

## Map controls

`add_map_controls()` adds fullscreen, globe-projection toggle, and scale bar
controls to any map.

```{r controls-default}
blank_map(projection = "none") |> add_map_controls()
```

Disable the globe toggle and move the scale bar:

```{r controls-custom}
grey_map() |>
    add_map_controls(globe = FALSE, scale_position = "bottom-right")
```

## Inline H3J source (requires duckdb, h3jsr, dplyr)

`add_h3j_source_inline()` encodes a data frame of H3 cells as a `data:` URI
and registers it as a vector source — no file or tile server needed. Use it
with any `mapgl` layer function (e.g. `mapgl::add_fill_layer()`).

```{r h3j-inline}
library(duckdb)
library(h3jsr)
library(dplyr)

con <- dbConnect(duckdb())

dbSendQuery(con, "install httpfs; load httpfs;")

df <- dbGetQuery(con, "
    SELECT records, cell
    FROM read_parquet('s3://obis-products/speciesgrids/h3_7/*')
    WHERE species = 'Minuca rapax'
")

df <- df |>
    mutate(h3 = h3jsr::get_parent(cell, 3)) |>
    group_by(h3) |>
    summarise(value = sum(records))

blank_map(projection = "mercator") |>
    add_h3j_source_inline("h3source", df = df, h3_col = "h3") |>
    mapgl::add_fill_layer(
        id           = "h3layer",
        source       = "h3source",
        fill_color   = mapgl::interpolate(
            column = "value",
            values = c(0, 5, 10, 15, 20),
            stops  = c("#2c7bb6", "#abd9e9", "#ffffbf", "#fdae61", "#d7191c")
        ),
        fill_opacity = 0.8
    ) |>
    mapgl::fit_bounds(get_species_extent(taxonid = 955271))
```

## DeckGL H3 widget — `maplibre_h3()`

`maplibre_h3()` creates a self-contained htmlwidget combining a MapLibre GL
basemap with a DeckGL `H3HexagonLayer`. Suitable for species richness, record
counts, or any per-H3-cell metric.

```{r h3-widget-flat}
# df from the duckdb query above
maplibre_h3(df, center = c(-50, -10), zoom = 2)
```

Enable 3-D extrusion with `extruded = TRUE`:

```{r h3-widget-3d}
maplibre_h3(df, center = c(-50, -10), zoom = 2, extruded = TRUE, elevation_scale = 1000)
```
