---
title: "Static Maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Static Maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(obisrecipes)
library(ggplot2)
```

The static-map functions build ggplot2 world maps layer by layer. The workflow
is:

1. Create a basemap with `blank_map_s()` or `grey_map_s()`.
2. Convert raw occurrence records to a spatial grid with one of the
   `occ_to_*grid()` helpers, or fetch pre-gridded data from OBIS.
3. Overlay the grid with `add_species_data_s()`.

## Basemaps

### Blank white map

```{r blank-default}
blank_map_s()
```

Keep the graticule for maps where coordinate reference is important:

```{r blank-graticule}
blank_map_s(keep_graticule = TRUE)
```

Any PROJ string can be passed to `map_crs`:

```{r blank-mollweide}
blank_map_s(map_crs = "+proj=moll")
```

### Grey OBIS-style map

`grey_map_s()` replicates the style used on OBIS taxon pages: grey ocean,
white land, and a semi-transparent GEBCO bathymetry overlay.

```{r grey-default}
grey_map_s()
```

Disable bathymetry when it is not needed:

```{r grey-no-bath}
grey_map_s(add_bathymetry = FALSE)
```

Regional views are controlled with `plot_xlim` / `plot_ylim`:

```{r grey-regional}
grey_map_s(plot_xlim = c(-80, 20), plot_ylim = c(10, 70))
```

## Grid converters (offline)

These functions convert a data frame of occurrence records (columns
`decimalLongitude` and `decimalLatitude`) into an sf polygon layer with one
polygon per occupied grid cell.

```{r mock-data}
occ_mock <- data.frame(
    decimalLongitude = c(-122.4, -118.2, -87.6, 2.3, 151.2, -43.1, 18.4, 103.8),
    decimalLatitude  = c(37.8,   34.1,   41.9,  48.9, -33.9,  -22.9, 59.3,  1.3)
)
```

### Geohash grid

```{r geohash}
occ_to_geohashgrid(occ_mock)
```

```{r geohash-fine}
occ_to_geohashgrid(occ_mock, grid_res = 4)
```

### H3 hexagonal grid

```{r h3}
occ_to_h3grid(occ_mock)
```

```{r h3-fine}
occ_to_h3grid(occ_mock, grid_res = 4)
```

### A5 pentagonal grid

```{r a5}
occ_to_a5grid(occ_mock)
```

```{r a5-fine}
occ_to_a5grid(occ_mock, grid_res = 4)
```

## Preparing grid data for plotting

`prepare_grid_data()` bins a numeric `records` column into ordered factor
levels matching the OBIS colour breaks, ready for `scale_fill_manual()`.

```{r prepare}
prepare_grid_data(occ_to_geohashgrid(occ_mock))
```

Custom break thresholds:

```{r prepare-custom}
prepare_grid_data(occ_to_geohashgrid(occ_mock), breaks = c(1, 5, 50, 500, 5000))
```

## Fetching data from OBIS (requires internet)

`get_occ_gridded()` retrieves raw occurrence records via the OBIS API and
converts them to a spatial grid. `get_occ_tiles()` uses the pre-aggregated
tile endpoint — much faster for large taxa.

```{r fetch-grid, eval = FALSE}
# Whale shark (taxonid 141438)
grid_geohash <- get_occ_gridded(taxonid = 141438, type_grid = "geohash", grid_res = 3)
grid_h3      <- get_occ_gridded(taxonid = 141438, type_grid = "h3",      grid_res = 3)
grid_a5      <- get_occ_gridded(taxonid = 141438, type_grid = "a5",      grid_res = 3)

# With date filter
grid_recent <- get_occ_gridded(
    taxonid   = 141438,
    startdate = "2010-01-01",
    enddate   = "2024-12-31",
    type_grid = "geohash",
    grid_res  = 3
)
```

```{r fetch-tiles, eval = FALSE}
tiles       <- get_occ_tiles(taxonid = 141438, grid_res = 3)
tiles_r4    <- get_occ_tiles(taxonid = 141438, grid_res = 4)
tiles_dated <- get_occ_tiles(
    taxonid   = 141438,
    startdate = "2020-01-01",
    enddate   = "2024-12-31",
    grid_res  = 3
)
```

## Plotting occurrence grids

`add_species_data_s()` overlays a spatial grid on an existing basemap. By
default it zooms to the bounding box of the data.

```{r plot-geohash, eval = FALSE}
blank_map_s() |>
    add_species_data_s(grid_data = grid_geohash)
```

Pass `auto_breaks = TRUE` to let ggplot2 choose break thresholds automatically:

```{r plot-h3-auto, eval = FALSE}
grey_map_s() |>
    add_species_data_s(
        grid_data   = grid_h3,
        auto_breaks = TRUE
    )
```

Global view without zooming to the data extent:

```{r plot-tiles-global, eval = FALSE}
grey_map_s() |>
    add_species_data_s(
        grid_data     = tiles,
        limit_by_bbox = FALSE
    )
```

Custom breaks, colours, and no legend:

```{r plot-custom, eval = FALSE}
blank_map_s() |>
    add_species_data_s(
        grid_data = tiles,
        breaks    = c(1, 5, 50, 500),
        colors    = c("#f7fbff", "#6baed6", "#2171b5", "#084594"),
        legend    = FALSE
    )
```

Regional view — Indo-Pacific:

```{r plot-regional, eval = FALSE}
grey_map_s(plot_xlim = c(30, 160), plot_ylim = c(-40, 35)) |>
    add_species_data_s(
        grid_data     = tiles,
        limit_by_bbox = FALSE,
        plot_xlim     = c(30, 160),
        plot_ylim     = c(-40, 35)
    )
```
